---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MALDIpqi

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/436219305.svg)](https://zenodo.org/badge/latestdoi/436219305)
<!-- badges: end -->

MALDIpqi calculates Parchment Glutamine Index from MALDI TOF ZooMS data.
It is a sample level measure of the glutamine deamidation from a list of peptides.

For details in the data processing and mathematical method and models to estimate PQI, refer to our publication Nair et al. (2022)

The method was initially developed to estimate parchment quality based on deamidation, following some ideas from Wilson et al. (2012). However, it can be applied in other tissues as in van Doorn et al. (2012) or Brown et al. (2021)

## Installation

You can install the released version of MALDIpqi from github with:

``` r
install.packages('devtools')
install.github("ismaRP/MALDIpqi")
```

## References
Nair, B. et al. (2022) ‘Parchment Glutamine Index (PQI): A novel method to estimate glutamine deamidation levels in parchment collagen obtained from low-quality MALDI-TOF data’, bioRxiv. doi:10.1101/2022.03.13.483627.

Wilson, J., van Doorn, N.L. and Collins, M.J. (2012) ‘Assessing the extent of bone degradation using glutamine deamidation in collagen’, Analytical chemistry, 84(21), pp. 9041–9048.

van Doorn, N.L. et al. (2012) ‘Site-specific deamidation of glutamine: a new marker of bone collagen deterioration’, Rapid communications in mass spectrometry: RCM, 26(19), pp. 2319–2327.

Brown, S. et al. (2021) ‘Examining collagen preservation through glutamine deamidation at Denisova Cave’, Journal of archaeological science, 133, p. 105454.

## Example

```{r load, eval=T}
data_folder = "data/mzML"
results_folder = "results_test"
```


This is minimal example of the workflow. It assumes the spectra are in `r data_folder`,
in mzML format and with file names in the form "samplename_replicate.ext". Where the replicate number is 1, 2 or 3 and ext is the extension of the files.

```{r example, eval=F}
library(MALDIpqi)

iso_peaks = getIsoPeaks(
  indir=data_folder, outdir=NULL, readf="mzml",
  peptides = peptides, chunks = 50, n_isopeaks = 5, min_isopeaks = 4,
  smooth_method = "SavitzkyGolay", hws_smooth = 8, halfWindowSize = 20, SNR = 1.5)

q2e = wls_q2e(peptides = peptides, n_isopeaks = 5,
              data_list = iso_peaks, outdir=results_folder)

res_free_gamma = lme_pqi(q2e, peptides=peptides, outdir=results_folder,
                         logq = T, n_isopeaks=5, g="free")
res_fixed_gamma = lme_pqi(q2e, peptides=peptides, outdir=results_folder,
                          logq = T, n_isopeaks=5, g=-1/2)

```

